A significance test in forward stepwise model selection

### 2013-12-05 paper/paper.tex
Chipping away

We allow forward stepwise selection to add groups of variables in each
step, not only in the case of binary encoding for categorical
variables but also for any grouping purpose. For example, groups of
variables may be pre-designated factors such as expression
measurements for all genes in a single functional pathway. An entire
group is included or excluded together from the final model. For
consistency we will use $g, h$ as indices rather than the usual $i, j$
throughout. Since single variables can be considered groups of size 1,
our general exposition includes non-grouped situations as a special case.

@@

We allow for the possibility $p > n$ in which case \ref{eq:gmodel} is
generically underdetermined. In such cases it still may be possible to
estimate $\beta$ well if it is sparse--that is, if it has few nonzero
entries \citep{donoho:pursuit}. In the rest of this paper we refer to
variable groups $X_g$ as noise variables if $\beta_g$ is a zero vector
and as true predictor variables if $\beta_g$ has nonzero entries.

Before we describe the forward stepwise procedure we require one last
ingredient. To each group of variables we assign a weight $w_g$. These
weights act like penalties or costs, so increasing $w_g$ makes it
more difficult for the group $X_g$ to enter the model. The
modeler can choose weights arbitrarily, but we will only use one
particular choice, based on $p_g$, that we discuss later. With this we
are ready to describe the forward stepwise procedure, given in Algorithm
\ref{algo:fs}. Note that to enable our p-value computations,
\textit{max.steps} should be at most $\min (n, G) - 1$. In our
implementation we treat the active set $A$ as an ordered set to easily
track the order of variables entering the model.

\begin{algorithm}
\DontPrintSemicolon
\KwData{An $n$ vector $y$ and $n \times p$ matrix $X$ of $G$ grouped variables}
\KwResult{Active set $A$ of variable groups included in the model}
\SetKwFunction{lsfitResidual}{lsfitResidual}
\caption{Forward stepwise procedure with groups and weights}
\BlankLine
$A \gets \emptyset$\;
$A^c \gets \{ 1, \ldots, G \}$\;
$r_0 \gets y$\;
\For{$step \leftarrow 1$ \KwTo $max.steps$}{
  $g^* \gets \argmax_{g \in A^c} \norm{X_g^T r_{g-1}}_2 / w_g$\;
  $A \gets A \cup \{ g^* \}$\;
  $A^c \gets A^c \backslash \{ g^* \}$\;
  \For{$g \in A^c$}{
    $X_g \gets (I - X_{g^*}X_{g^*}^\dagger) X_g$;
  }
  $r_g \gets $\lsfitResidual{$r_{g-1}, X_{g^*}$}\;
}
\KwRet{$A$}
\label{algo:fs}
\end{algorithm}

Subset selection refers to the problem of choosing a subset of the $p$
potential variables to include in a model. Since there are $2^p$ such
choices, exhaustive search is computationally infeasible when $p$ is
large. If exhaustive search were possible, performing a large
search runs the risk of over-fitting unless model complexity is
appropriately penalized. Algorithm \ref{algo:fs} produces a much
smaller set of potential models, at most \textit{max.steps}. However,
it is a greedy algorithm so the set of models it produces may not
contain the best possible model. This is an inherent shortcoming which
we do not address in this paper.

@@

Although forward stepwise is a greedy algorithm producing a
potentially sub-optimal sequence of models, under favorable conditions
it can still perform well. There is a parallel literature to the
compressed sensing literature \textit{(cite some here)} with theorems
stating that forward stepwise can exactly select the true model under
some stringent conditions involving quantities like the sparsity of
the true model and the coherence of the design matrix. We do not
pursue this any further because we find empirically that forward
stepwise can work well in much less restrictive situations than those
guaranteed by theorems.

To explore the performance of forward stepwise empirically we
conducted simulations with results shown in ~\ref{fig:fwdstepsims}

\begin{figure}
\begin{center}
\subfigure[Gaussian design]{
%\label{fig:smallgauss}
\includegraphics[width=0.5\textwidth]{../figs/fwdstep/gaussian_p2n_n100_lower1p4_upper1p6.pdf}}
\hspace{-15pt}
\subfigure[Categorical design]{
%\label{fig:smallcat}
\includegraphics[width=0.5\textwidth]{../figs/fwdstep/gaussian_p2n_n100_lower1p4_upper1p6_corr0p2.pdf}}
\caption{\small \it The left panel shows results of a simulation with
  an independent Gaussian design matrix and a signal vector having
  three groups with non-zero coefficients of sizes one, two, and
  three, and seven groups of noise variables with sizes varying from
  one to ten, for a total of 30 variables. For the right panel we
  repeated this setup, but group sizes corresponded to levels of
  categorical variables. The groups of size one were increased to two,
so there are 10 categorical variables with a total of 32 levels.}
\label{fig:fwdstepsim}
\end{center}
\end{figure}

@@

\cite{significance:lasso} prove that under the null hypothesis that $A_k$
contains all the strong predictor variables, the distribution of a
test statistic $T_k \propto \lambda_k(\lambda_k - \lambda_{k+1})$ is
asymptotically Exp(1). In the lasso case we know a lot about the knots
and active set, but the group lasso picture is slightly more
complicated. For the group lasso, $\hat \beta(\lambda)$ does not have
piecewise linear components. To overcome this difficulty we will
restrict our attention to the first group of variables to enter the
active set since the analysis then follows almost exactly as for the
lasso. (\todo Change this if I can find $\lambda_k$).

The \emph{group lasso estimator} is the following solution to a convex
problem
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\begin{aligned}
\label{eq:gsoln}
\displaystyle \hat \beta_\lambda = \argmin_{\beta \in \real^p} \frac{1}{2} \| y - X \beta \|_2^2 +
   \lambda \sum_{g=1}^G w_g \| \beta_g \|_2
\end{aligned}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The parameter $\lambda \geq 0$ enforces sparsity in groups: for large
$\lambda$ most of the $\beta_g$ will be zero vectors. The weights
$w_g$ are usually taken to be $\sqrt p_g$ to normalize the penalty
across groups.  Note that this includes the usual lasso estimator as a
special case when all of the groups are of size 1, since then the
penalty term is the $L^1$-norm of $\beta$.

>>>>

We allow forward stepwise selection to add groups of variables in each
step, not only in the case of binary encoding for categorical
variables but also for any grouping purpose. For example, groups of
variables may be pre-designated factors such as expression
measurements for all genes in a single functional pathway. An entire
group is included or excluded together from the final model. For
consistency we will use $g, h$ as indices rather than the usual $i, j$
throughout. Since single variables can be considered groups of size 1,
our general exposition includes non-grouped situations as a special
case. Our variant of forward stepwise uses a different method than
usual for choosing which group to add because we are using the
same objective in the forward stepwise procedure that was used to
derive our test statistic.

@@

We allow for the possibility $p > n$ in which case ~\ref{eq:gmodel} is
generically underdetermined. In such cases it still may be possible to
estimate $\beta$ well if it is sparse--that is, if it has few nonzero
entries \citep{donoho:pursuit}. In the rest of this paper we refer to
variable groups $X_g$ as noise variables if $\beta_g$ is a zero vector
and as true predictor variables if $\beta_g$ has nonzero entries.

Before we describe the forward stepwise procedure we require one last
ingredient. To each group of variables we assign a weight $w_g$. These
weights act like penalties or costs, so increasing $w_g$ makes it
more difficult for the group $X_g$ to enter the model. The
modeler can choose weights arbitrarily, but we will only use one
particular choice, based on $p_g$, that we discuss later. With this we
are ready to describe the forward stepwise procedure, given in Algorithm
\ref{algo:fs}. Note that to enable our p-value computations,
\textit{max.steps} should be at most $\min (n, G) - 1$. If the noise
variance is known--as we assume unless specified otherwise--and
$\beta$ is expected to be quite sparse, then \textit{max.steps} can be
much smaller as long as it is still above the expected sparsity of
$\beta$. This last constraint is suggested so that forward stepwise
still has a chance of recovering all the nonzero coefficients of
$\beta$ before reaching \textit{max.steps}. In our implementation we
treat the active set $A$ as an ordered set to easily track the order
of variables entering the model.

\begin{algorithm}
\DontPrintSemicolon
\KwData{An $n$ vector $y$ and $n \times p$ matrix $X$ of $G$ grouped variables}
\KwResult{Active set $A$ of variable groups included in the model}
\SetKwFunction{lsfitResidual}{lsfitResidual}
\caption{Forward stepwise procedure with groups and weights}
\BlankLine
$A \gets \emptyset$\;
$A^c \gets \{ 1, \ldots, G \}$\;
$r_0 \gets y$\;
\For{$s \gets 1$ \KwTo $max.steps$}{
  $g^* \gets \argmax_{g \in A^c} \{ \norm{X_g^T r_{s-1}}_2 / w_g \}$\;
  $P \gets I - X_{g^*}X_{g^*}^\dagger$\;
  $A \gets A \cup \{ g^* \}$\;
  $A^c \gets A^c \backslash \{ g^* \}$\;
  \For{$g \in A^c$}{
    $X_g \gets P X_g$\;
  }
  $r_s \gets P r_{s-1}$\;
}
\KwRet{$A$}
\label{algo:fs}
\end{algorithm}

Subset selection refers to the problem of choosing a subset of the $G$
potential groups to include in a model. Since there are $2^G$ such
choices, exhaustive search is computationally infeasible when $G$ is
large. If exhaustive search were possible, performing a large
search runs the risk of over-fitting unless model complexity is
appropriately penalized. Forward stepwise produces a much
smaller set of potential models, with cardinality at most \textit{max.steps}. However,
it is a greedy algorithm so the set of models it produces may not
contain the best possible model. This is an inherent shortcoming of
forward stepwise procedures and should be kept in mind when choosing
between model selection methods.

@@

Although forward stepwise is a greedy algorithm producing a
potentially sub-optimal sequence of models, under favorable conditions
it can still perform well. There is a parallel literature to the
compressed sensing literature \textit{(cite some here)} with theorems
stating that forward stepwise can exactly select the true model under
some stringent conditions involving quantities like the sparsity of
the true model and the coherence of the design matrix. Recall that the
coherence $\mu(X)$ of a matrix $X$ with columns scaled to have unit
2-norm is defined as

\begin{equation}
  \mu(X) = \max_{i \neq j} \{ | \innerp{ X_i }{ X_j } | \}
\end{equation}

Writing $k$ for the sparsity of $\beta$, typical results in this
literature say that if $k < (1/\mu + 1)/2$ and the nonzero
coefficients of $\beta$ are sufficiently large then forward stepwise
recovers $\beta$ with high probability. We refer the
reader to the literature for details. For our purposes the
conditions required to guarantee exact recovery are usually far too
stringent. We conduct simulations to show empirically that forward
stepwise can work well even when it is not working perfectly, and that
it does so under a much wider range of conditions.

\begin{figure}
\begin{center}
\subfigure[Ternary design]{
\label{fig:fsternary}
\includegraphics[width=0.5\textwidth]{../figs/fwdstep/ternary_p2n_n100_lower1p4_upper1p4.pdf}}
\hspace{-15pt}
\subfigure[Gaussian design with correlation]{
\label{fig:fsgausscorr}
\includegraphics[width=0.5\textwidth]{../figs/fwdstep/gaussian_p2n_n100_lower1p4_upper1p6_corr0p2.pdf}}
\caption{\small \it The left panel shows the results of the simulation
  using design matrices with i.i.d. ternary entries taking values $0,
  \pm 1$ with equal probability. The left panel shows the results
  with design matrices of standard Gaussian entries with
  equi-correlation of 0.2. }
\label{fig:fwdstepsim}
\end{center}
\end{figure}

For various sparsity levels $k$, Algorithm ~\ref{algo:fs} was run on
various data sets with max.steps equal to $k$ and the ``average
power'' was calculated as the number of true variables in the active
set after $k$ steps divided by $k$. Nonzero coefficients had magnitude
in a range of multiples of $\sqrt{2\log(p)}$, e.g. in $[1.4
\sqrt{2\log(p)}, 1.6 \sqrt{2\log(p)}]$. Results are shown in
Figure~\ref{fig:fwdstepsim}. Some calculations show the required
sparsity level to guarantee exact recovery in these situations is
about 2 or smaller, and the required nonzero coefficient magnitude is
likely in the range of 10 to 100 times $\sqrt{2\log(p)}$.

@@

\cite{significance:lasso} prove that under the null hypothesis that $A_k$
contains all the strong predictor variables, the distribution of a
test statistic $T_k \propto \lambda_k(\lambda_k - \lambda_{k+1})$ is
asymptotically Exp(1). In the lasso case we know a lot about the knots
and active set, but the group lasso picture is slightly more
complicated. For the group lasso, $\hat \beta(\lambda)$ does not have
piecewise linear components. To overcome this difficulty we will
restrict our attention to the first group of variables to enter the
active set since the analysis then follows almost exactly as for the
lasso.

The \emph{group lasso estimator} is a solution to the following convex
problem
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{equation}
\begin{aligned}
\label{eq:gsoln}
\displaystyle \hat \beta_\lambda = \argmin_{\beta \in \real^p} \frac{1}{2} \| y - X \beta \|_2^2 +
   \lambda \sum_{g=1}^G w_g \| \beta_g \|_2
\end{aligned}
\end{equation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The parameter $\lambda \geq 0$ enforces sparsity in groups: for large
$\lambda$ most of the $\beta_g$ will be zero vectors. The weights
$w_g$ are usually taken to be $\sqrt p_g$ to normalize the penalty
across groups.  Note that this includes the usual lasso estimator as a
special case when all of the groups are of size 1, since then the
penalty term is the $L^1$-norm of $\beta$.
